pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-07 manuscript/greenwell.tex
started working on reviewers comments

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

@@

Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of two or more features as well.

@@

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops. This makes it easy to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package. To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load parallel backend
registerDoParallel(cores = 4)  # use 4 cores
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The result is displayed in Figure~\ref{fig:partial_par}.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, .parallel = TRUE)
\end{example}

@@

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. %Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, .progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

library(ggplot2)
ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
  geom_tile() +
  geom_contour(color = "white", alpha = 0.5) +
  scale_fill_distiller(palette = "Spectral") +
  theme_bw() +
  facet_grid(~ Species)
\end{example}

@@

>>>>

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes $E\left[\boldsymbol{\mathcal{Y}}\right] = \boldsymbol{X}\boldsymbol{\beta}$). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be
more easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

@@

Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of two or more features as well.

@@

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops. This makes it easy to request progress bars (e.g., \code{progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package. To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load parallel backend
registerDoParallel(cores = 4)  # use 4 cores
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The result is displayed in Figure~\ref{fig:partial_par}.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE)
\end{example}

@@

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. %Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

library(ggplot2)
ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
  geom_tile() +
  geom_contour(color = "white", alpha = 0.5) +
  scale_fill_distiller(name = "Average\nlogit", palette = "Spectral") +
  theme_bw() +
  facet_grid(~ Species)
\end{example}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Using partial with caret
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
